---
title: "CSCI S-89B Final Project – STM Topic Modeling on News Headlines"
author: "InJoo Kim"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(stringr)
  library(quanteda)
  library(stm)
})

options(stringsAsFactors = FALSE)

# Resolve project root whether you run from repo root or from /notebooks
project_root <- if (dir.exists("data")) {
  getwd()
} else if (dir.exists("../data")) {
  normalizePath("..")
} else {
  getwd()
}

path_sample <- file.path(project_root, "data", "sample", "news_headlines_sample.csv")
path_figs   <- file.path(project_root, "report", "figures")
path_tables <- file.path(project_root, "report", "tables")
if (!dir.exists(path_figs)) dir.create(path_figs, recursive = TRUE)
if (!dir.exists(path_tables)) dir.create(path_tables, recursive = TRUE)
```

# 1. Goal

I use **Structural Topic Modeling (STM)** to discover latent themes in a large collection of news headlines, and I use **metadata (publish date)** to see how topic prevalence changes over time.

# 2. Load sample data

```{r}
if (!file.exists(path_sample)) {
  stop(paste0(
    "Sample file not found at: ", path_sample, "\n",
    "Create it with: Rscript src/01_make_sample.R <raw_csv> data/sample/news_headlines_sample.csv 20000"
  ))
}

news <- readr::read_csv(path_sample, show_col_types = FALSE)

dplyr::glimpse(news)
news %>% dplyr::slice_head(n = 5)
```

# 3. Basic cleaning + date parsing

```{r}
news_clean <- news %>%
  transmute(
    headline = as.character(headline_text),
    publish_date_raw = as.character(publish_date)
  ) %>%
  filter(!is.na(headline), str_length(str_trim(headline)) > 0) %>%
  mutate(
    # ABC News date format is YYYYMMDD (e.g., 20171212)
    publish_date = suppressWarnings(ymd(publish_date_raw, tz = "UTC")),
    publish_date = as.Date(publish_date)
  ) %>%
  filter(!is.na(publish_date)) %>%
  mutate(
    year = lubridate::year(publish_date)
  )

summary(news_clean$publish_date)
```

# 4. Tokenization + DTM (quanteda → stm format)

```{r}
# Minimal preprocessing for headlines: lowercasing + keep letters/spaces
text_clean <- news_clean$headline %>%
  str_to_lower() %>%
  str_replace_all("[^a-z\\s]", " ") %>%
  str_squish()

toks <- tokens(text_clean, remove_punct = TRUE, remove_numbers = TRUE)
toks_clean <- tokens_remove(toks, stopwords("en"))

# drop very short tokens
toks_clean <- tokens_keep(toks_clean, pattern = "^[a-z]{3,}$", valuetype = "regex")

dfm_mat <- dfm(toks_clean)

# Remove very rare terms to stabilize STM
dfm_mat <- dfm_trim(dfm_mat, min_termfreq = 5)

# Remove documents with no features (empty after preprocessing)
non_empty <- rowSums(dfm_mat) > 0
dfm_mat <- dfm_mat[non_empty, ]
meta <- news_clean[non_empty, ]

stm_in <- convert(dfm_mat, to = "stm")
docs  <- stm_in$documents
vocab <- stm_in$vocab

list("n_docs" = length(docs), "vocab_size" = length(vocab), "meta_rows" = nrow(meta))
```

# 5. Pick K (searchK) + Fit final STM

```{r}
set.seed(123)

K_grid <- c(5, 10, 15, 20)

k_search <- searchK(
  documents  = docs,
  vocab      = vocab,
  K          = K_grid,
  prevalence = ~ year,
  data       = meta
)

plot(k_search)

# Choose K 
K_final <- 10

stm_model <- stm(
  documents  = docs,
  vocab      = vocab,
  data       = meta,
  K          = K_final,
  prevalence = ~ year,
  verbose    = TRUE
)
```

# 6. Inspect topics

```{r}
labelTopics(stm_model, n = 10)

# Save summary plot using base R graphics
png(filename = file.path(path_figs, paste0("topic_summary_K", K_final, ".png")),
    width = 1200, height = 800, res = 120)
plot(stm_model, type = "summary", main = "Topic Prevalence (K=10)")
dev.off()

# Display in notebook
plot(stm_model, type = "summary", main = "Topic Prevalence (K=10)")
```

# 7. Topic prevalence over time (estimateEffect)

```{r}
eff <- estimateEffect(1:K_final ~ year, stmobj = stm_model, metadata = meta)

# Save prevalence plot using base R graphics
png(filename = file.path(path_figs, paste0("topic_prevalence_over_time_K", K_final, ".png")),
    width = 1200, height = 800, res = 120)
plot(
  eff,
  covariate   = "year",
  method      = "continuous",
  topics      = c(1, 2, 3),
  model       = stm_model,
  printlegend = TRUE,
  main        = "Topic Prevalence Over Time"
)
dev.off()

# Display in notebook
plot(
  eff,
  covariate   = "year",
  method      = "continuous",
  topics      = c(1, 2, 3),
  model       = stm_model,
  printlegend = TRUE,
  main        = "Topic Prevalence Over Time"
)
```

# 8. Word clouds (topic-level)

```{r}
# Wordcloud is base plotting; save via png()
if (!requireNamespace("wordcloud", quietly = TRUE)) {
  install.packages("wordcloud")
}
library(wordcloud)

for (t in c(1, 2, 3)) {
  png(filename = file.path(path_figs, paste0("topic_wordcloud_topic", t, "_K", K_final, ".png")),
      width = 900, height = 900)
  cloud(stm_model, topic = t, scale = c(4, 0.5))
  dev.off()
}
```

# 9. Topic quality (semantic coherence / exclusivity)

```{r}
# Calculate semantic coherence and exclusivity separately
semcoh <- semanticCoherence(stm_model, docs)
excl <- exclusivity(stm_model)

q_df <- data.frame(
  topic = 1:K_final,
  semantic_coherence = semcoh,
  exclusivity = excl
)
print(q_df)
readr::write_csv(q_df, file.path(project_root, "report", "tables", paste0("topic_quality_K", K_final, ".csv")))
```

